---
title: "Ravienne_UROP_Zoes_Code"
author: "Zoe Habel"
date: "r Sys.Date()"
output:
  github_document: default
  html_notebook: default
---
#### Loading packages

```{r Packages, echo=T, message=FALSE}
library(tidyverse) 
```


*tidyverse includes dplyr*

#### Here's where I loaded my data locally

```{r Raw Data, echo = T, results = 'hide'}
load("~/Desktop/git/urop_ravienne/ICPSR UROP Data/ICPSR_21600 4/DS0005/21600-0005-Data.rda")
load("~/Desktop/git/urop_ravienne/ICPSR UROP Data/ICPSR_21600/DS0001/21600-0001-Data.rda")
load("~/Desktop/git/urop_ravienne/ICPSR UROP Data/ICPSR_21600 8/DS0022/21600-0022-Data.rda")
```

#### [google sheet link :)](https://docs.google.com/spreadsheets/d/1Qvp9XRvgWT0kIqNywmpVKmovxEQQrm3WWV7K6oD7JFQ/edit?gid=0#gid=0)

## **Zoe's code as of 3/19/2025**

### Step 1: selecting the variables from each wave and organizing them

```{r}
#wave 1 variables
data_wave1 <- da21600.0001 |> 
  select(
    AID, H1TO12, H1TO17
  )
#wave 2 variables
data_wave2 <- da21600.0005 |> 
  select(
    AID, H2ED19, H2ED17, H2PR1, H2PR2, H2PR3, H2PR4
  )
#wave 4 variables
data_wave4 <- da21600.0022 |>
  select(
    AID,
    H4MA1,
    H4MA3,
    H4MA5,
    H4TO51,
    H4TO58,
    H4TO46,
    H4TO48,
    H4TO52,
    H4TO47,
    H4TO53,
    H4TO54,
    H4TO55,
    H4TO56,
    H4TO59,
    H4TO60,
    H4TO61,
    H4TO62,
    #lifetime alc use more than 2 or three times
    H4TO33,
  )
```

# 33. Have you had a drink of beer, wine, or liquor more than two or three times? Do not include sips or tastes from someone else's drink

if else yes - H4TO33 -> skip all other alc questions
legit skip # : 3322 

# 35. During the past 12 months, on how many days did you drink alcohol?

H4TO35 -> if 'none' - skip H4TO36:H4TO39 
legit skip # : 3322  


# 36. Think of all the times you have had a drink during the past **12 months.** How many drinks did you usually have each time? A 'drink' is a glass of wine, a can or bottle of beer, a wine cooler, a shot glass of liquor, or a mixed drink.

H4TO36 -> skip H4TO37:H4TO42T  
legit skip # : 4427
+ 1105 skipped (H4TO35 + H4TO33)

# 39. During the past 30 days, on how many days did you drink?

H4TO39 -> if 'none' - skip H4TO40:H4TO42T  
legit skip # :  4427 

# 40. Think of all the times you have had a drink during the past **30 days.** How many drinks did you usually have each time? A 'drink' is a glass of wine, a can or bottle of beer, a wine cooler, a shot glass of liquor, or a mixed drink.

H4TO40 ->  skip H4TO41:H4TO42T
legit skip # : 6307 
+ 1880 skipped

(10 more than 'no' trying to figure out why)

H4TO42H 
legit skip # : 12445
+ 6138 skipped (skipped from 33 + 39 + 40)

# 43. Was there ever a period in your life when you drank more alcohol than you do now?
H4TO43 -> if 'no' skip H4TO45:H4TO46
legit skip #: 3322 (skipped from H4TO33)

# 44. During the period when you drank the most, on how many days did you drink?
H4TO44 -> 
legit skip #: 8723 (skipped from 33 + 43)
+ 5401 skipped 
(23 more than 'no')

# 45. During the period when you drank the most, how many drinks did you usually have each time?
legit skip #: 8723 (skipped from 33 + 43)

if else criteria met -> skip all rest

# 46. How often has your drinking interfered with your responsibilities at work or school?

H4TO46 ->  
legit skip #: 9330

+ 607 skipped 
of 607 -- 
193 refused/don't know or not asked on pretest
414 did not meet inclusion criteria

H4TO47	
# How often have you been under the influence of alcohol when you could have gotten yourself or others hurt, or put yourself or others at risk, including unprotected sex?
legit skip #: 9330

H4TO48	
# How often have you had legal problems because of your drinking, like being arrested for disturbing the peace or driving under the influence of alcohol, or anything else?
legit skip #: 9330

H4TO49
# 49. How often have you had problems with your family, friends, or people at work or school because of your drinking?
if 'never' -> skip 51
legit skip #: 9330

H4TO50
# 50. Did you continue to drink after you realized drinking was causing you problems with family, friends, or people at work or school?
legit skip #: 14011 
+ 4681 skipped ('never' 49 + skip 45)

H4TO51	
# Have you ever found that you had to drink more than you used to in order to get the effect you wanted?
legit skip #: 9330 (skip 45)
other missing: 5

H4TO52	
# Has there ever been a period when you spent a lot of time drinking, planning how you would get alcohol, or recovering from a hangover?
other missing: 8

H4TO53	
# Have you often had more to drink or kept drinking for a longer period of time than you intended?
other missing: 7

H4TO54	
# Have you ever tried to quit or cut down on your drinking?
-> if 'yes' - skip H4TO55 (3594)
-> if else - skip H4TO56 (no = 2765)
other missing: 12

H4TO55	
# 55. Has there ever been a period of time when you wanted to quit or cut down on your drinking?
legit skip #:12924
+ 3594 skipped ('yes' 55)
other missing: 11

H4TO56	
# 56. When you decided to cut down or quit drinking, were you able to do so for at least one month?
legit skip #: 12103
+ 2773 skipped (if else 'yes' 54)
if 'yes' -> skip H4TO57
other missing: 5

H4TO57
# 57. How many times have you tried but been unable to cut down or quit drinking for at least one month?
legit skip #: 15434
+ 3331 skipped ('yes' H4TO56)
other missing: 7

H4TO58	
# During the first few hours of not drinking do you experience withdrawal symptoms such as the shakes, feeling anxious, trouble getting to sleep or staying asleep, nausea, vomiting, or rapid heartbeats?
legit skip #: 9330 (skip from 45)
other missing : 9

H4TO59	
# Have you ever continued to drink after you realized drinking was causing you any emotional problems (such as feeling irritable, depressed, or uninterested in things or having strange ideas) or causing you any health problems (such as ulcers, numbness in your hands/feet or memory problems)?
legit skip #: 9330 (skip from 45)
other missing : 12
H4TO60	
# Have you ever given up or cut down on important activities that would interfere with drinking like getting together with friends or relatives, going to work or school, participating in sports, or anything else?
legit skip #: 9330 (skip from 45)


(Q51 = 1) 5
+ (Q52 = 1) 8
+ (Q53 = 1) 7
+ (Q56 = 0) 5
+ (Q55 = 1) 11
+ (Q58 = 1) 7
+ (Q59 = 1) 12
+ (Q60 = 1) 9
all missing not skipped : 64

If drinking experiences >= 3 then display 'You said that you...'
...
Ask Q.61.
Else -> skip to Q.63.

If drinking experiences > 3, insert 'at least three of.'

Other missing drinking exp: 
# 61. Did (at least three of) these experiences occur together in a 12-month period?
legit skip #: 13629
+ 4299
might include NAs

-- 
# 62. 
legit skip #: 14300
+ 671 ('no' 61)

### Step 2: merging the variables by AID

```{r}
data_waves1and2 <- merge(
  data_wave1, 
  data_wave2, 
  by = "AID"
)
data_waves124 <- merge(
  data_waves1and2,
  data_wave4,
  by = "AID"
)
```

### Step 3: Using the dplyr function mutate to combine variables from subvariables e.g. ACEs

#### ACES

```{r}
data_waves124$emotional_abuse <- as.numeric(data_waves124$H4MA1 == "(5) (5) More than ten times")
data_waves124$physical_abuse <- as.numeric(
  data_waves124$H4MA3 == "(3) (3) Three to five times" |
    data_waves124$H4MA3 == "(4) (4) Six to ten times" |
    data_waves124$H4MA3 == "(5) (5) More than ten times"
)
data_waves124$sexual_abuse <- as.numeric(data_waves124$H4MA5 != "(6) (6) This has never happened")
data_waves124 <- data_waves124 |>
  mutate(aces = emotional_abuse + physical_abuse + sexual_abuse)
```

### Step 3.2: Creating composites for trickier variables

#### perceived discriminiation & social support


### Step 3.2: Creating composites for trickier variables 
#### perceived discriminiation & social support

```{r}
#perceived discriminiation
data_waves124$teacher_discrimination <- data_waves124$H2ED19 |>
  fct_rev() |>
  as.numeric()

data_waves124$peer_prejudice <- data_waves124$H2ED17 |>
  fct_rev() |>
  as.numeric()

data_waves124 <- data_waves124 |> mutate(
  (perceived_discrimination = replace_na(teacher_discrimination, 0) +
               replace_na(peer_prejudice, 0)))

#perceived social support

data_waves124$adult_support <- data_waves124$H2PR1 |>
  fct_rev() |>
  as.numeric()
data_waves124$teacher_support <- data_waves124$H2PR2 |>
  fct_rev() |>
  as.numeric()
data_waves124$parent_support <- data_waves124$H2PR3 |>
  fct_rev() |>
  as.numeric()
data_waves124$peer_support <- data_waves124$H2PR4 |>
  fct_rev() |>
  as.numeric()

data_waves124 <- data_waves124 |>
  mutate(perceived_social_support =
    replace_na(adult_support, 0) + replace_na(teacher_support, 0)
    + replace_na(parent_support, 0) + replace_na(peer_support, 0)
  )
```

### Step 3.3: Cleaning + creating composites for questions related to problematic alcohol use


#### Domain approach + study skip logic

```{r}
#early life substance use
data_waves124$early_life_alc_use <- as.numeric(data_waves124$H1TO12 == "(1) (1) Yes")

data_waves124$early_life_heavy_drinking <- data_waves124$H1TO17 |>
  fct_rev() |>
  as.numeric()

#alcohol use wave 4

data_waves124$lifetime_alc_use <- as.numeric(data_waves124$H4TO33 == "(1) (1) Yes")


data_waves124 <- data_waves124 |>
  mutate(
    early_life_subst_use = replace_na(early_life_alc_use, 0) + replace_na(early_life_heavy_drinking, 0)
  )
#Have you ever found that you had to drink more than you used to in order to get the effect you wanted?
data_waves124$sud_progression1 <- as.numeric(data_waves124$H4TO51 == "(1) (1) Yes")

#During the first few hours of not drinking do you experience withdrawal symptoms such as the shakes, feeling anxious, trouble getting to sleep or staying asleep, nausea, vomiting, or rapid heartbeats?
data_waves124$sud_progression2 <- as.numeric(data_waves124$H4TO58 == "(1) (1) Yes")

#How often has your drinking ever interfered with your responsibilities at work or school?
data_waves124$sud_progression3 <- data_waves124$H4TO46 |>
fct_collapse("(1) Once or more"= c("(1) (1) 1 time","(2) (2) More than 1 time"))
data_waves124$sud_progression3 <-as.numeric(
  data_waves124$sud_progression3=="(1) Once or more")

#How often have you had legal problems because of your drinking, like being arrested for disturbing the peace or driving under the influence of alcohol, or anything else?
data_waves124$sud_preoccupation1 <- data_waves124$H4TO48 |>
fct_collapse("(1) Once or more"= c("(1) (1) 1 time","(2) (2) More than 1 time"))


data_waves124$sud_preoccupation1 <-as.numeric(
  data_waves124$sud_preoccupation1=="(1) Once or more")

#Has there ever been a period when you spent a lot of time drinking, planning how you would get alcohol, or recovering from a hangover?
data_waves124$sud_preoccupation2 <- as.numeric(data_waves124$H4TO52 == "(1) (1) Yes")

#How often have you been under the influence of alcohol when you could have gotten yourself or others hurt, or put yourself or others at risk, including unprotected sex?
data_waves124$sud_preoccupation3 <- data_waves124$H4TO47 |>
fct_collapse("(1) Once or more"= c("(1) (1) 1 time","(2) (2) More than 1 time"))
data_waves124$sud_preoccupation3 <-as.numeric(
  data_waves124$sud_preoccupation1=="(1) Once or more")

# Have you often had more to drink or kept drinking for a longer period of time than you intended?
data_waves124$sud_losscontrol1 <- as.numeric(data_waves124$H4TO53 == "(1) (1) Yes")

#Has there ever been a period of time when you wanted to quit or cut down on your drinking?
data_waves124$sud_losscontrol2 <- as.numeric(data_waves124$H4TO55 == "(1) (1) Yes")

#Have you ever tried to quit or cut down on your drinking?
data_waves124$sud_losscontrol3 <- as.numeric(data_waves124$H4TO54 == "(1) (1) Yes")

#When you decided to cut down or quit drinking, were you able to do so for at least one month?
data_waves124$sud_losscontrol4 <- data_waves124$H4TO56 |>
  fct_rev() |>
  as.numeric() 

#Have you ever continued to drink after you realized drinking was causing you any emotional problems (such as feeling irritable, depressed, or uninterested in things or having strange ideas) or causing you any health problems (such as ulcers, numbness in your hands/feet or memory problems)?

data_waves124$sud_persistence1 <- as.numeric(data_waves124$H4TO59 == "(1) (1) Yes")

#Have you ever given up or cut down on important activities that would interfere with drinking like getting together with friends or relatives, going to work or school, participating in sports, or anything else?

data_waves124$sud_persistence2 <- as.numeric(data_waves124$H4TO60 == "(1) (1) Yes")

# Create composite domains with NAs replaced by 0 for calculation

data_waves124 <- data_waves124 |>
  mutate(
    sud_progression = replace_na(sud_progression1, 0) + replace_na(sud_progression2, 0) + replace_na(sud_progression3, 0),
    sud_preoccupation = replace_na(sud_preoccupation1, 0) + replace_na(sud_preoccupation2, 0) + replace_na(sud_preoccupation3, 0),
    sud_losscontrol = replace_na(sud_losscontrol1, 0) + replace_na(sud_losscontrol2, 0) + replace_na(sud_losscontrol3, 0) + replace_na(sud_losscontrol4, 0),
    sud_persistence = replace_na(sud_persistence1, 0) + replace_na(sud_persistence2, 0)
  )

# Create composite drinking experiences
data_waves124 <- data_waves124 |>
  mutate(
    drinking_exp = sud_progression + sud_preoccupation + sud_losscontrol + sud_persistence
  )

data_waves124 <- data_waves124 |>
  mutate(substance_misuse = if_else(H4TO61 == "(1) (1) Yes" &
                                      H4TO62 >= 18, 1, 0))
#here's where we could change the criteria - "drinking experiences >= 3" is what the study used as criteria to ask Q61, that could be replaced with other criteria and potentially leave out Q61

data_waves124 <- data_waves124 |>
  mutate(
    substance_misuse_manual = if_else(
      drinking_exp >= 3 & H4TO61 == "(1) (1) Yes" & H4TO62 >= 18, 
      1,
      0))

AUD_indicators <- data_waves124 |>
  select(AID,
         substance_misuse_manual,
         substance_misuse,
         drinking_exp,
         starts_with("sud"),
         H4TO47:H4TO62,
        )
write_csv(x = AUD_indicators, file = "problematic_alcohol_misuse_addhealth.csv")
```

### Step 4: Check that the composites were created successfully

```{r,eval=F,include=T}
addhealth_clean_and_composites <- (data_waves124[, c(1, 27:60)])
summary(addhealth_clean_and_composites)
write_csv(x=addhealth_clean_and_composites,"addhealth_clean_and_composites")
```


## Missing Data handling

```{r}
#which(is.na(AUD_indicators$sud_progression1))
```

## **Ravienne’s code as of 03/09/25**

```{r,eval=F,include=T}
#selecting the variables from each wave and organizing them
library(dplyr)
w1vars <- da21600.0001 %>% select(AID, H1TO12, H1TO17)
w2vars <- da21600.0005 %>% select(AID, H2ED19, H2ED17, H2PR1, H2PR2, H2PR3, H2PR4)
w4vars <- da21600.0022 %>% select(
  AID,
  H4MA1,
  H4MA3,
  H4MA5,
  H4TO51,
  H4TO58,
  H4TO46,
  H4TO48,
  H4TO52,
  H4TO47,
  H4TO53,
  H4TO54,
  H4TO55,
  H4TO56,
  H4TO59,
  H4TO60,
  H4TO61,
  H4TO62
)

#merging the variables by AID
merg1 <- merge(w1vars, w2vars, by = "AID")
merg2 <- merge(merg1, w4vars, by = "AID")

#use dplyr function mutate to combine variables from subvariables e.g. ACEs

#aces
merg2$emotional_abuse <- as.numeric(merg2$H4MA1 == "(5) (5) More than ten times")
merg2$physical_abuse <- as.numeric(
  merg2$H4MA3 == "(3) (3) Three to five times" |
    merg2$H4MA3 == "(4) (4) Six to ten times" |
    merg2$H4MA3 == "(5) (5) More than ten times"
)
merg2$sexual_abuse <- as.numeric(merg2$H4MA5 != "(6) (6) This has never happened")
merg2 <- merg2 %>% mutate(aces = emotional_abuse + physical_abuse + sexual_abuse)

#early life substance use
merg2$early_life_alc_use <- as.numeric(merg2$H1TO12 == "(1) (1) Yes")
merg2$early_life_heavy_drinking <- levels(merg2$H1TO17) <- c(6, 5, 4, 3, 2, 1, 0)
merg2 <- merg2 %>% mutate(early_life_subst_use = early_life_alc_use + early_life_heavy_drinking)

#problematic substance use - create a manual diagnosis variable and send dr liu the codebooks


#perceived discriminiation
merg2$teacher_discrimination <- levels(merg2$H2ED19) = c(1, 2, 3, 4, 5)
merg2$peer_prejudice <- levels(merg2$H2ED17) = c(5, 4, 3, 2, 1)
merg2 <- merg2 %>% mutate(perceived_discrimination = teacher_discrimination + peer_prejudice)

#perceived social support
merg2$adult_support <- levels(merg2$H2PR1) = c(1, 2, 3, 4, 5)
merg2$teacher_support <- levels(merg2$H2PR2) = c(1, 2, 3, 4, 5)
merg2$parent_support <- levels(merg2$H2PR3) = c(1, 2, 3, 4, 5)
merg2$peer_support <- levels(merg2$H2PR4) = c(1, 2, 3, 4, 5)
merg2 <- merg2 %>% mutate(perceived_social_support = adult_support + teacher_support + parent_support + peer_support)
```

